---
title: "DESeq2_RNAseq"
author: "Cèsar-Lluís"
date: "04/11/2025"
output: html_document
knit: (function(input, ...) {
  knitr::opts_knit$set(root.dir = '/Users/csierra/mount/point2/CRG/Naive/RNA_N/wo18')
  rmarkdown::render(input, ...)
})
---
#Set working directory
```{r setup}
knitr::opts_knit$set(root.dir = normalizePath('/Users/csierra/mount/point2/CRG/Naive/RNA_N/wo18'))
knitr::opts_chunk$set(echo = TRUE)
setwd(knitr::opts_knit$get("root.dir"))
# getwd()
```
#Load libraries
```{r}
lapply(c("DESeq2", "pheatmap", "ggplot2","data.table", "readxl",
         "biomaRt", "dplyr", "ggrepel", "writexl", "geneset", 
         "genekitr", "ChIPseeker", "forcats"),
       library, character.only = T, verbose = F)
```

#My Counts
```{r}
# List all featureCounts output files
files <- list.files(path = "counts", pattern = "\\.featureCounts$",
  full.names = TRUE)

# Read and merge all files by GeneID
count_list <- lapply(files, function(file) {
  dat <- fread(file, skip = 1)
  sample_name <- gsub("\\.featureCounts.*", "", basename(file))
  dat <- dat[, .(Geneid, Count = dat[[ncol(dat)]])]
  setnames(dat, "Count", sample_name)
  return(dat)
})

# Merge all by Geneid
merged_counts <- Reduce(function(x, y) merge(x, y, by = "Geneid", all = TRUE), count_list)

# Set Geneid as rownames
merged_counts <- as.data.frame(merged_counts)
rownames(merged_counts) <- merged_counts$Geneid
merged_counts$Geneid <- NULL

# Save to a single file
#fwrite(merged_counts, "counts/combined_gene_counts.txt", sep = "\t", quote = FALSE)
```


#Load raw counts into DESeq objects
```{r}
getwd()
sampletable <- read.table("sample_sheet.txt", header=T, sep="\t")
rownames(sampletable) <- sampletable$Sample_Name

se_star_matrix <- DESeqDataSetFromMatrix(countData = merged_counts,
                                  colData = sampletable,
                                  design = ~ Type)
```

#Filtering
```{r}
nrow(se_star_matrix)
se_star_matrix <- estimateSizeFactors(se_star_matrix)
se_star_matrix <- se_star_matrix[rowSums(counts(se_star_matrix)) > 10, ]
nrow(se_star_matrix)
#save this as universe
universe<-rownames(se_star_matrix)
universe<-as.data.frame(universe)
#Remove version from gene name
universe$universe <- gsub("\\.\\d+$", "", universe$universe)
#write_xlsx(universe, "tables/universe.xlsx")

```

#Fit statistical model

```{r}
se_star2 <- DESeq(se_star_matrix)
```

##Save the normalized counts for further usage
```{r}
tx2gene <- read.table("tx2gene.gencode.m38.csv")

# compute normalized counts (log2 transformed); + 1 is a count added to avoid errors during the log2 transformation: log2(0) gives an infinite number, but log2(1) is 0.
# normalized = TRUE: divide the counts by the size factors calculated by the DESeq function
norm_counts <- log2(counts(se_star2, normalized = TRUE)+1)

# write normalized counts to text file
#write.table(norm_counts, "normalized_counts.txt", quote=F, col.names=T, row.names=T, sep="\t")
```

#Visualization
## Transform raw counts to be able to visualize the data.
DESeq2 developper advice to use: rlog (Regularized log) or vst (Variance Stabilizing Transformation) transformations for visualization and other applications other than differential testing:
VST runs faster than rlog. If the library size of the samples and therefore their size factors vary widely, the rlog transformation is more robust.
Both options produce log2 scale data which has been normalized by the DESeq2 method with respect to library size.

```{r}
#rlog transformation:
rld <- rlog(se_star2)
#Remove batch effect?
#assay(rld) <- limma::removeBatchEffect(assay(rld), rld$Batch)

```

## Samples correlation

```{r}
# calculate between-sample distance matrix
sampleDistMatrix <- as.matrix(dist(t(assay(rld))))

pheatmap(sampleDistMatrix)

#PCA
plot<-plotPCA(object = rld,
		intgroup = "Type", ntop=2500)
plot + geom_label(aes(label = name))

# ggsave(filename = "plots/PCA.pdf",
#       plot = plot,
#      bg = "transparent",
#     width = 4, height = 4, units = "in", dpi = 300)
```
#WTN vs TSN
```{r}
deseq2.resWTN_TSN <- lfcShrink(se_star2, contrast=c("Type","TSN","WTN"), type="ashr")
deseq2.res.dfWTN_TSN <- as.data.frame(deseq2.resWTN_TSN)
deseq2.res.dfWTN_TSN$genes <- row.names(deseq2.res.dfWTN_TSN)
```

#Check triplicated genes
```{r}
#Get gene names
ensembl<-useMart("ENSEMBL_MART_MOUSE")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")
ensembl=useDataset("mmc57bl6nj_gene_ensembl", mart=ensembl)

#Remove version from gene name
deseq2.res.dfWTN_TSN$genes <- gsub("\\.\\d+$", "", deseq2.res.dfWTN_TSN$genes)

genes<-deseq2.res.dfWTN_TSN$genes

df<-getBM(attributes = c("external_gene_name","ensembl_gene_id"),
      filters= "ensembl_gene_id",
      values=genes,
      mart=mouse)
colnames(df)<-c("external_gene_name", "genes")

dfnames<-left_join(deseq2.res.dfWTN_TSN, df, by="genes")

#Include triplicated info
triplicated<-read.csv("../genes_triplicated.csv")
colnames(triplicated)<-c("chromosome", "external_gene_name", "HSA")
dfnames_tr<-left_join(dfnames, triplicated, by="external_gene_name",relationship = "many-to-many")

dfnames_tr <- dfnames_tr %>%  mutate(is.na = is.na(chromosome))
dfnames_tr <- dfnames_tr %>% mutate(triplicated = ifelse(is.na==c("FALSE"), "Yes", "No"))
```

#save degs tables
```{r}
WTN_TSN_sign<-subset(dfnames_tr, padj<0.05)

WTN_TSN_sign<-WTN_TSN_sign %>% mutate(DE=case_when(
  log2FoldChange > 0 ~ "Up",
  log2FoldChange < 0 ~ "Down"
))

#save table
# write_xlsx(WTN_TSN_sign, "tables/DEGs_WTN_TSN.xlsx")
# write_xlsx(dfnames_tr, "tables/DEGs_WTN_TSN_all.xlsx")

```


#Count DEGs and plot triplicated and non-triplicated
```{r}
ggplot(WTN_TSN_sign, aes(x = triplicated, fill = DE)) +
  geom_bar() +
  labs(title = "Number of Genes in Each Region by Direction and Triplicated Status",
       x = "Region",
       y = "Number of Genes") +
  theme_minimal()

# ggsave(filename = "plots/barplot_triplicated.pdf",
#        bg = "transparent", 
#        width = 7, height = 5, units = "in", dpi = 300)

#Normalize by size

tblup<-subset(WTN_TSN_sign, DE=="Up")
tbldown<-subset(WTN_TSN_sign, DE=="Down")

chrup<-lapply(tblup, table)
chrcountup<-chrup$triplicated
chrcountup<-as.data.frame(chrcountup)
chrcountup$Freq<-as.numeric(chrcountup$Freq)
chrcountup$Direction<-"Up"

chrdown<-lapply(tbldown, table)
chrcountdown<-chrdown$triplicated
chrcountdown<-as.data.frame(chrcountdown)
chrcountdown$Freq<-as.numeric(chrcountdown$Freq)
chrcountdown$Direction<-"Down"

chrcount<-rbind(chrcountup, chrcountdown)

genenum<-c(24178, 305, 24178)
chrcount$DE<-(chrcount$Freq/genenum)*100

ggplot(chrcount, aes(x=Var1, y=DE)) + geom_histogram(stat="identity", xlab="Chromosome", aes(fill=Direction)) + theme(panel.background = element_blank(),  axis.line = element_line(colour = "black"))+ scale_y_continuous(expand = expansion(mult = c(0, 0)))

# ggsave(filename = "plots/barplot_triplicated_norm.pdf",
#        bg = "transparent", 
#        width = 7, height = 5, units = "in", dpi = 300)

```

#Chromosome distribution
```{r}
ensembl<-useMart("ENSEMBL_MART_MOUSE", host = "useast.ensembl.org")
ensembl=useDataset("mmc57bl6nj_gene_ensembl", mart=ensembl)
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")

alldeg<-WTN_TSN_sign
genes<-unique(alldeg[c("genes")])

table<-getBM(attributes = c("chromosome_name","ensembl_gene_id","external_gene_name"),
      filters= "ensembl_gene_id",
      values=genes,
      mart=mouse)


table$chromosome_name <- factor(table$chromosome_name,levels = c("1", "2", "3", "4","5", "6", "7", "8","9", "10", "11", "12","13", "14", "15", "16","17", "18", "19", "X"))

colnames(table)<-c("chromosome_name", "ensembl_gene_id", "genes")

names(alldeg)[names(alldeg) == 'genes'] <- 'ensembl_gene_id'

tbl<-full_join(table, alldeg, by = "ensembl_gene_id", copy = FALSE)
#I had to manually add some chromosomes that were not included by getBM OR erase the ones with NA
#tbl2<-tbl[1:335,]

ggplot(data=tbl, aes(x=chromosome_name)) + geom_histogram(stat="count", aes(fill=DE)) +
  theme(panel.background = element_blank(),  axis.line = element_line(colour = "black"))

tblup<-subset(tbl, DE=="Up")
tbldown<-subset(tbl, DE=="Down")


chrup<-lapply(tblup, table)
chrcountup<-chrup$chromosome_name
chrcountup<-as.data.frame(chrcountup)
chrcountup$Freq<-as.numeric(chrcountup$Freq)
chrcountup$Direction<-"Up"


chrdown<-lapply(tbldown, table)
chrcountdown<-chrdown$chromosome_name
chrcountdown<-as.data.frame(chrcountdown)
chrcountdown$Freq<-as.numeric(chrcountdown$Freq)
chrcountdown$Direction<-"Down"

chrcount<-rbind(chrcountup, chrcountdown)

library(TxDb.Mmusculus.UCSC.mm10.knownGene)
genenum <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)
genenum<-as.data.frame(genenum)
genenum<-table(unlist(genenum$seqnames))

chrcount$genenum<-genenum[1:20]
chrcount$DE<-(chrcount$Freq/chrcount$genenum)*100
ggplot(chrcount, aes(x=Var1, y=DE)) + geom_histogram(stat="identity", xlab="Chromosome", aes(fill=Direction)) + theme(panel.background = element_blank(),  axis.line = element_line(colour = "black"))+ scale_y_continuous(expand = expansion(mult = c(0, 0)))

ggsave(filename = "plots/barplot_chr.pdf",
       bg = "transparent", 
       width = 7, height = 5, units = "in", dpi = 300)
```



#Volcanoplot
```{r}
WTN_TSN_sign$logp <- -log10(WTN_TSN_sign$padj)
with(WTN_TSN_sign, plot(log2FoldChange, logp,  pch=20, main="Volcano plot", xlim=c(-2,2)))
with(subset(WTN_TSN_sign, padj<.05 ), points(log2FoldChange, logp, pch=20, col="red"))

volcano = ggplot(WTN_TSN_sign, aes(x = log2FoldChange, y = logp, color = triplicated)) +
  geom_point(alpha = 0.2, size = 1.5) +
  scale_y_continuous(trans = "log1p") +
  theme_bw(base_size = 14) +
  
  # subset of genes to label
  geom_segment(
    data = subset(WTN_TSN_sign,
                  (logp > 6 & (log2FoldChange > 1 | log2FoldChange < -1)) |
                    external_gene_name %in% c("Son","Dyrk1a","Synj1","App","Brwd1")),
    aes(x = log2FoldChange,
        y = logp,
        xend = log2FoldChange,
        yend = logp + 0.5),
    linewidth = 0.3,
    color = "grey30"
  ) +
  
  geom_text_repel(
    data = subset(WTN_TSN_sign,
                  (logp > 6 & (log2FoldChange > 1 | log2FoldChange < -1)) |
                    external_gene_name %in% c("Son","Dyrk1a","Synj1","App","Brwd1")),
    aes(label = external_gene_name),
    size = 3,
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.3, "lines")
  ) +
  
  ylab("-log10(p.adj)") +
  xlab("Log Fold Change") +
#  xlim(-2.1, 3) +
  theme(legend.position = "none") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(-0.2, 0.2),
             linetype = "dashed") 


volcano
ggsave(filename = "plots/volcanoWTN_TSN.pdf",
       plot = volcano,
       bg = "transparent",
       width = 5, height = 5, units = "in", dpi = 300)
```


#Check  degree of overexpression of triplicated genes
```{r}
WTN_TSN_sign$FC<-2^WTN_TSN_sign$log2FoldChange
WTN_TSN_sign_up<-subset(WTN_TSN_sign, DE=="Up")

data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
 return(data_sum)
}

df2 <- data_summary(WTN_TSN_sign_up, varname="FC", 
                    groupnames=c("triplicated"))
# Convert dose to a factor variable
df2$triplicated=as.factor(df2$triplicated)
head(df2)

ggplot(df2, aes(x=triplicated, y=FC)) + geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=FC-sd, ymax=FC+sd), width=.2,
                 position=position_dodge(.9)) + scale_y_continuous(expand = expansion(mult = c(0, 0)))
```

#GSEA
## Get gene set
```{r}
gs <- geneset::getGO(org = "mouse",ont =c("bp"))
#ALT
#load("mouse_H_v5.RData")
```

## Transform to geneList
```{r}
#1
DEGs<-as.data.frame(dfnames_tr)
geneList = as.numeric(DEGs[,2])
names(geneList)= row.names(DEGs) 
geneList_1 = sort(geneList, decreasing = TRUE)
```

#GSEA analysis
```{r}
gse_1 <- genGSEA(genelist = geneList_1, geneset = gs, q_cutoff = 1,p_cutoff = 1, min_gset_size = 10)
gse_1_df<-as.data.frame(gse_1$gsea_df) #ns
```

#Plot shinyGO
```{r}
shiny<-read.csv("shinygo/GOBP_all.csv", header = T)
shiny_filt<-shiny[c(9,11,6, 4, 8, 5),]
shiny_filt$GeneRatio<-shiny_filt$nGenes/677

shiny_filt %>%
  mutate(Pathway = fct_reorder(Pathway, nGenes)) %>%
  ggplot(aes(x = GeneRatio, y = Pathway, color = Enrichment.FDR)) +
  
  # Draw the "lollipop stick"
  geom_segment(aes(x = 0, xend = GeneRatio, y = Pathway, yend = Pathway)) +
  
  # Draw the "lollipop head"
  geom_point(aes(size = nGenes)) +
  
  # Color gradient
scale_color_gradient(low = "red2", high = "mediumblue", limits = c(0, 0.1)) +
  
  theme_classic() +
  labs(x = "Gene Ratio", y = "Pathway") +
  theme(
    legend.position = "right"
  )
ggsave("plots/GOBP_all.pdf", height = 3, width = 6.5, dpi=600)
```

